Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-19 paper/paper.tex
Fix duplicate lambda
Add text in intro about section 3

The goal of this paper is to give a tutorial on proof by logical relation for
dependent types in a simple and digestable format. \scw{This sentence should come as 
early as possible.}
The system we present is small but covers
important features\scw{list these features explicitly: 
large eliminations, indexed types, others? \ldots}  that are only available or require special treatments in
dependently typed systems. We choose boolean
types over natural numbers as our base type for simplicity, but include an
intensional identity type in our type system to show how indexed
types are treated in a logical relational proof. \scw{Talk about 
identity type first.} Unlike
\citet{nbeincoq,decagda,martin-lof-a-la-coq} but similar to
\citet{anand2014towards}, we include an infinite hierarchy of
universes to not only support type-level computations, but also avoid
the unnecessary code duplication pointed out by \citet{nbeincoq} when
big and small types are treated non-uniformly.

@@

In Section~\ref{sec:extension}, \scw{what about sections 2-4?}

 we extend our logical
relation to prove the stronger property that every well-typed term
has a normal form. Most interestingly, we find that extending the
language with
$\eta$ rule for functions and the metatheoretic result to normalization for open terms are
mostly orthogonal to the fact that the language is dependently
typed. In our mechanized proof, the overall
structure of the logical relation remains unchanged, and the additional
proof obligations are mostly related to properties about the untyped
lambda calculus, all of which can be derived through syntactic
means.
In other words, once we have established the base technique for
proof by logical relation for dependent types, we can factor out the
complexities of an extension the are not specifically related to dependent
types; such an extension, if desired, may be studied in the context of
a simply typed language and later ported into a dependently typed setting.

@@

\section{Toward Decidability of Type Conversion and $\eta$ laws}
\label{sec:extension}
\begin{figure}[h]
  \[
    \begin{array}{lcl}
      % \beta\text{-}\mathit{neutral\ terms}\\
      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
                    f]] \\ \\
      % \beta\text{-}\mathit{normal\ terms} \\
      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
                    |\ [[f ~ f : f]]\\
            & |   & [[\\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
    \end{array}
  \]
  \caption{$\beta$-neutral and normal forms}
  \label{fig:nenf}
\end{figure}
\begin{figure}[h]
  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
  \caption{Extended logical relation}
  \label{fig:logrelopen}
\end{figure}

>>>>

The goal of this paper is to give a tutorial on proof by logical relation for
dependent types in a simple and digestable format. \scw{This sentence should come as
early as possible.} Dependently typed systems comes in many flavors and
vary greatly in their expressiveness.
In Section~\ref{sec:spec}, we introduce \lang{}, a dependently typed
language that is small but relatively complete in features \scw{list these features explicitly:
large eliminations, indexed types, others? \ldots} related to
dependently types, including large elimination, a full universe
hierarchy, an intensional identity type and a boolean base type, both
of which are equipped with dependent elimination forms (e.g. J eliminator
for the identity type). In particular, the inclusion of identity
types, which is absent from \citet{decagda}, \citet{nbeincoq},
helps demonstrate how indexed types are handled when defining the
logical relation.

% We choose boolean
% types over natural numbers as our base type for simplicity, but include an
% intensional identity type in our type system to show how indexed
% types are treated in a logical relational proof. \scw{Talk about
% identity type first.} Unlike
% \citet{nbeincoq,decagda,martin-lof-a-la-coq} but similar to
% \citet{anand2014towards}, we include an infinite hierarchy of
% universes to not only support type-level computations, but also avoid
% the unnecessary code duplication pointed out by \citet{nbeincoq} when
% big and small types are treated non-uniformly.
In Section~\ref{sec:logreldep}, we give the definition of our logical
relation as an inductively defined relation. We first show that the
logical relation respects evaluation (referred to as the irrelevance
property in \citet{martin-lof-a-la-coq}), then proceed to
prove that the logical relation corresponds to a partial function in
the sense that the each type has a unique interpretation if there
exists one.

In Section~\ref{sec:logrelproof}, we use the standard technique that
is commonly seen in ...

@@

In Section~\ref{sec:extension}, \scw{what about sections 2-4?}
 we extend our logical
relation to prove the stronger property that every well-typed term
has a normal form. Most interestingly, we find that extending the
language with
$\eta$ rule for functions and the metatheoretic result to normalization for open terms are
mostly orthogonal to the fact that the language is dependently
typed. In our mechanized proof, the overall
structure of the logical relation remains unchanged, and the additional
proof obligations are mostly related to properties about the untyped
lambda calculus, all of which can be derived through syntactic
means.
In other words, once we have established the base technique for
proof by logical relation for dependent types, we can factor out the
complexities of an extension the are not specifically related to dependent
types; such an extension, if desired, may be studied in the context of
a simply typed language and later ported into a dependently typed setting.

@@

\section{Toward Decidability of Type Conversion and $\eta$ laws}
\label{sec:extension}
\begin{figure}[h]
  \[
    \begin{array}{lcl}
      % \beta\text{-}\mathit{neutral\ terms}\\
      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
                    f]] \\ \\
      % \beta\text{-}\mathit{normal\ terms} \\
      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
                    |\ [[f ~ f : f]]\\
            & |   & [[\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
    \end{array}
  \]
  \caption{$\beta$-neutral and normal forms}
  \label{fig:nenf}
\end{figure}
\begin{figure}[h]
  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
  \caption{Extended logical relation}
  \label{fig:logrelopen}
\end{figure}
